Investigation of ant cuticle dataset using image texture analysis

### 2022-08-20 paper/main.tex
fix fig 7
address comments

\title{Investigation of ant cuticle dataset using image texture
    analysis methods}

@@

\address{%
    \addr{\affilnum{1}}{Laboratory of Machine Vision and Security Research, %
        College of Computing and Software Engineering, Kennesaw State %
        University, Marietta GA, USA}%
    \addr{\affilnum{2}}{College of Science and Mathematics, Kennesaw State
        University, Kennesaw, GA, USA}%
    \addr{\affilnum{3}}{Department of Mathematics, Kennesaw State University,
        Kennesaw, GA, USA}%
    \addr{\affilnum{4}}{Department of Mathematics, Science, and Informatics,
        Mercer University, Atlanta, GA, USA}%
}
\corraddr{chung1@kennesaw.edu}

@@

\begin{abstract}
    Ant cuticle texture presumably provides some type of function, and therefore
    is useful to research for ecological applications and bioinspired designs.
    In this study, we employ statistical image texture analysis and deep machine
    learning methods to automatically group similar ant species based on
    morphological features. We establish a public database of ant cuticle images
    for research. We provide a comparative study of the performance of image
    texture classification and deep machine learning methods on ant cuticle
    images. Our results show that the deep learning methods are giving higher
    accuracy in recognizing ant cuticle textures. Our experiments also reveal
    that the deep learning networks designed for image texture performs better
    than the general deep learning networks.
\end{abstract}
\keywords{texture analysis; image processing; classification; machine learning;
    ant cuticle images; ecology}
\maketitle

@@

Insects compose half of biodiversity and rank among the most dominant organisms
in terrestrial ecosystems \cite{sheikh_diverse_2017}. A key factor for the
ecological success of insects is their exoskeleton, also known as cuticle. The
cuticle protects insects from predation, provides structural support, prevents
desiccation, and serves as a canvas for advertising visual and chemical signals
\cite{gullan_insects_2009}. Research has focused on the macrostructures and
internal chemical components that describe the functionality of the exoskeleton
\cite{gunderson_insect_1989}. More recent work is being done to understand the
functional aspects of external cuticle micro sculpturing
\cite{muthukrishnan_insect_2020, watson_diversity_2017}.

Due to the extensive number of insect species, manual exploration of
insect-based information is difficult and often requires expert training.
Therefore, automated entomology has attracted biologists and computer
scientists, and is expected to be a major contribution to the future of
insect-based research \cite{martineau_survey_2017}. One of the most commonly
used data types for insect analysis is image data. To develop an image-based
system for insect analysis, we can take advantage of existing work in general
image analysis methods.

We examine ants (\textit{Formicidae}) as they display an extreme diversity of
cuticle micro sculpturing across all subfamilies. Microsculpture ranges from
parallel longitudinal ridges to deep oval impressions to erratic protuberances.
Microsculpture has arisen convergently and independently throughout ant's
evolutionary history, which suggests  that these are complex traits undergoing
selection. Cuticle microsculpture on ants may help increase strength and
rigidness, resist abrasion, increase internal and external surface area, resist
microbial growth, and rear beneficial anti-biotic producing bacteria
\cite{johnson_effect_2011, bruckner_relationship_2017, currie_coevolved_2006}.
These specific functions may be associated with certain sculpturing types. In
order to analyze those functions, it is necessary to segment the image and group
similar textures for further analysis.

There is a large variety of ant species and most species are diverse in terms of
size, shape, behaviors, and cuticle textures \cite{harris_glossary_1979}.
Based on our initial observation of the cuticle sculpturing in ants, it appears
that some type of textural structures can be identified in images. Therefore,
this image dataset may be worth to be explored from the image texture analysis
point of view for the further knowledge discovery. Image texture analysis has
been used in image processing for many decades \cite{hung_image_2019}. Due to its
diversity of images on scales, smoothness, coarseness, microtexture, and
macrotexture occurring in images, it is a difficult problem for image textural
interpreters. With a rapid development of deep machine learning, deep neural
networks have been frequently used in image texture analysis
\cite{sun_robust_2014,liu_bow_2019,liu_pestnet_2019}.

Our contribution is to establish an ant image dataset which can be used to
explore the categorization of different ant cuticle textures using image texture
analysis methods. We share how the dataset of ant cuticle images is established
in Section 3. Our experiment shows that the deep learning networks designed for
image texture analysis give a better accuracy than the K-views clustering
method.

This paper is organized as below:  Section 2 reviews the microsculpturing
identification in ants, and image texture analysis and insect classification.
Section 3 describes the dataset preparation and image texture analysis methods
used in our experiments. Section 4 gives the experimental results, visualization
of both raw and processed images and analysis. The conclusion then follows.

@@

One challenge for functional investigations of cuticle microsculpturing is the
complexity of terms used to describe these patterns. Complex terminology may be
useful for describing fine differences between closely related species, but
broader comparisons require a simplified classification scheme. We recently
developed a classification system for microsculpturing that consists of only
five categories: smooth, striate, punctate, reticulate, and tuberous
\cite{hellenbrand_ant_2022}. Using these categories to classify 11,722 ant
species, we found that the earliest ants were likely smooth and that cuticle
microsculpturing independently evolved later in multiple lineages
\cite{hellenbrand_ant_2022}. An even simpler classification scheme is to divide
ants into only two categories: smooth and rough. Because microsculptuing has
evolved numerous times in ants, this binary classification scheme is sufficient
to detect conditions that select for any sculpturing at all compared to a smooth
state.

Using either a simplified or binary classification scheme, the next challenge is
figuring out how to apply either of these schemes to large numbers of species.
We live in an era of big data, and insect researchers have access to millions of
images available through digital museum collections \cite{beaman_mass_2012}.
Going through these digital collections manually is not feasible for large
projects. Focusing only on ants, for example, there are over 14,000 described
species, and each species has multiple worker and reproductive castes that may
vary in their microsculpturing patterns. Even within an individual ant,
microsculpturing typically varies among and within body segments, which
exponentially increases the number of images that must be assessed. Applied
computing offers a solution to this problem through developing tools that
automate trait classification. Automated approaches would allow researchers to
leverage existing image repositories to extract large trait datasets that can
then be used to detect evolutionary patterns.

@@

Image texture analysis is an important part of image interpretation as many
objects appearing in the textural patterns either partially or completely inside
an image. This phenomenon appears frequently in many images we encountered in
our environment as almost everybody is using the iPhone to capture the photos.
Scientists are also studying their subjects depending on taking the images
remotely such as satellite images. With an abundant number of textural images
surrounding us, it is essential to have a reliable automatic tool for the
textural image interpretation.

Computer methods are developed for textural images based on the spatial
relationship of gray levels of pixels encoded in the texture
\cite{haralick_image_1985,hung_image_2019}. Those methods by de-facto follow the
steps used in the statistical pattern recognition
\cite{fukunaga_introduction_2013}. Therefore, textural features are extracted,
and the classification and segmentation methods are used for the interpretation
either using the supervised or unsupervised approaches. The interpretation
method is usually taken and modified from the traditional pattern recognition
methods. For example, the gray-level co-occurrence matrix (GLCM) and local
binary patterns (LBP) are two of methods for extracting features
\cite{haralick_textural_1973, goos_gray_2000}.

Similar to the traditional pattern recognition, classical image texture analysis
methods can be grouped into four categories: statistical, structural,
model-based, and transform-based methods \cite{bharati_image_2004}. Among all of
the categories, the statistical method is often used by employing the method
from the pattern recognition principle.  In doing so, an image texture is
represented as feature vector using feature extraction methods and then fed into
the algorithm. Markov random field models are also studied for textural image
interpretation \cite{hassner_use_1981, cross_markov_1983}. The transform-based
methods use some functions to decompose an image texture into a set of basic
feature images. Gabor filters and Wavelet expansions are two of the widely used
approaches \cite{bovik_multichannel_1990}.

While most feature vectors are established based on a single pixel, Liu et al.
gave a comprehensive survey on textural characterization in which they call this
feature extraction as texture representation \cite{liu_bow_2019}. Their survey
on Bag of Words and Convolutional Neural Networks (CNNs)
\cite{krizhevsky_imagenet_2017} are the spatial relationship concepts. The
K-views was developed for taking the spatial information in the clustering
approach \cite{hung_image_2019}.

In the past decade, deep learning using CNNs has emerged as the mainstream
technology for image analysis. Following this trend, various CNN-based network
architectures have been designed to specifically characterize textural images.
Among these, the Fisher-vector CNN descriptor (FV-CNN) proposed by Cimpoi et
al. \cite{cimpoi_deep_2015} is widely accepted as one of the most important
pioneering works. It applies FV pooling to deep features obtained via a CNN
pre-trained using the ImageNet \cite{krizhevsky_imagenet_2017} to obtain encoded features
for texture classification. Although it is capable of achieving much improved
classification accuracy comparing to traditional hand-crafted texture features,
FV-CNN does not support an end-to-end learning, where feature extraction,
encoding, and classification are separated from each other. To achieve an
end-to-end learning, Zhang et al. \cite{zhang_deep_2017} proposed the deep texture
encoding network (DeepTEN), in which a novel texture encoding layer is added to
a standard CNN architecture. Then, Xue et al. \cite{xue_deep_2018} constructed the
deep encoding pooling network (DEP), which improves over DeepTEN by integrating
local spatial characteristics into the texture representation. Based on DeepTEN
and DEP, Hu et al. \cite{hu_multi_2019} further developed the multi-level texture
encoding and representation (MuLTER) network, which embeds a learnable encoding
module at each convolutional layer so that encoding is performed for both
low-level and high-level features, yielding a multi-level texture
representation.

@@

\subsection{Insect classification}

@@

Feng et al. \cite{feng_automated_2013} apply an automated system to classify
moth images based on semantic related visual attributes, which are defined as a
pattern on the moth wings. Feng et al. \cite{feng_automated_2013} use a custom
texture descriptor based on the combination of GLCM and \textit{scale-invariant
    feature transform} (SIFT) features \cite{gotlieb_texture_1990,
    lowe_distinctive_2004}. The method proposed by Feng et al.
\cite{feng_automated_2013} is used to classify 50 different moth species across
8 families \cite{feng_automated_2013}. The results from Feng et al.
\cite{feng_automated_2013} suggest that traditional feature extraction
techniques for the semantic visual attributes of the moth wings are sufficient
for training a classifier to classify an image between 10 randomly selected moth
species.

Urteaga et al. \cite{urteaga_scorpions_2016} use machine learning methods in
order to classify images between two different scorpion species:
\textit{Centruroides limpidus} and \textit{Centruroides noxius}. After applying
background distinction based on dynamic color threshold, Urteaga et al.
\cite{urteaga_scorpions_2016} apply feature extraction to extract features from
the separated scorpion image such as aspect ratio, rectangularity, and
compactness. Urteaga et al. \cite{urteaga_scorpions_2016} apply three different
classification models to classify the image as one of the species: Artificial
Neural Network, Regression Tree, and Random Forest classifiers
\cite{urteaga_scorpions_2016}. The results from Urteaga et al.
\cite{urteaga_scorpions_2016} show that after background removal,
characteristics from the entire body of the scorpion can be used to create a
binary classifier that can classify the image as one of the two species.

@@

In this section, we describe the creation of the custom dataset used in this
research. In our dataset, we use ant head images from AntWeb
\cite{perrichot_antweb_2012} and define two categories for them based on the
appearance of the cuticle texture: \textit{rough} and \textit{smooth}. We also
refer to these ant head images, which describe the pose of the ant, as ant
cuticule images, due to the clear cuticle texture present on the head. Some
randomly selected images from each category are shown in Figures
\ref{fig:smooth-cuticle-texture} and \ref{fig:rough-cuticle-texture}.

@@

Initially, the sculpture identification protocol had 5 subcategories of cuticle
texture: smooth, punctate, striate, reticulate, and tuberous. For simplicity, we
work only with the two main categories. The training set identifications were
reviewed together as a group by the assistants. Once training was complete,
assistants were assigned the same genera of ants to identify independently each
week. A weekly meeting was held to discuss identifications and assign new ones.
These identifications were collected in the master spreadsheet and the
identifications were assigned to individual ant species on a majority basis.

To collect the images, the assistants followed the taxonomy information
available in the master spreadsheet to the appropriate
AntWeb\footnote{\url{https://www.antweb.org/}} page. In many cases, there are
multiple ant head images of the same species, and occasionally there are
multiple image resolution available from a single image. To simplify the data
collection process, the assistants were instructed to download the first ant
head image of the species being identified in the highest resolution possible.
Each image was named with an identifier that corresponds with the row number in
the master spreadsheet. The same ant head images that were downloaded in the
data collection phase were the same ones used in the sculpture identification
protocol. Ant species which did not have any images of the head were excluded
from the dataset. Additionally, ant species which only had a head image of a
queen ant were excluded from the dataset.

Ant specimen images taken from AntWeb \cite{perrichot_antweb_2012} are created
by different photographers and therefore have different attributes, such as
environment, resolution, and lighting. In the ant head images, the ant head is
in the center of the image and the body is pointing away from the camera. The
focus of the ant head image is centered on the head, with the background and
image artifacts from the ant body typically blurred. In most ant head images,
there is a bar which indicates the scale of the image due to the variety in the
sizes of different ant species. In a few ant head images, there exists some text
denoting the specimen identifier and other information. In terms of texture,
some ant specimens are very old, so their head images have other abnormalities
such as cracks in the cuticle and the presence of dust.

The completed dataset of contains 2,499 images. 1072 samples of rough textured
ant cuticle textures comprise 43\% of the dataset. The remaining 1427 samples of
smooth textured ant cuticle textures comprise 57\% of the dataset. Due to the
variety of the ant head image attributes, we apply simple preprocessing before
the images are used in our model. We want the images to have a uniform size for
simplicity in our classification process. Since the ant head images are
typically centered in the image, we apply a center crop to each image to create
a square image of the same size. Once the image is square, we resize each image
to a fixed size of 256x256 pixels. We leave other discrepancies in the images
untouched. A summary of the resulting dataset is shared in Table
\ref{tab:dataset-summary}.

@@

\newcommand{\segmentedsubwidth}{0.24\textwidth}
\begin{figure}
    \centering
    \begin{subfigure}{\segmentedsubwidth}
        \includegraphics[width=1\linewidth]{/home/ngardn10/projects/cuticulus/paper/assets/images/kviews/r103.png}
        \caption*{Rough texture}
    \end{subfigure}
    \begin{subfigure}{\segmentedsubwidth}
        \includegraphics[width=1\linewidth]{/home/ngardn10/projects/cuticulus/paper/assets/images/kviews/r111.png}
        \caption*{Smooth texture}
    \end{subfigure}
    \begin{subfigure}{\segmentedsubwidth}
        \includegraphics[width=1\linewidth]{/home/ngardn10/projects/cuticulus/paper/assets/images/kviews/s105.png}
        \caption*{Rough texture}
    \end{subfigure}
    \begin{subfigure}{\segmentedsubwidth}
        \includegraphics[width=1\linewidth]{/home/ngardn10/projects/cuticulus/paper/assets/images/kviews/s107.png}
        \caption*{Smooth texture}
    \end{subfigure}

@@

\FloatBarrier
\section{Conclusion}
Ant cuticle texture presumably has some function, but without the proper tools,
evaluating the function based on thousands of species is infeasible. We have
shown in this work that a deep learning approach and classical image texture
analysis methods can be used to automatically categorize ants based on their
cuticle texture, therefore supporting research on the evaluation of the function
in future work. Our categorization system is novel in the field of automated
insect identification due to the broad number of species captured by it.
Additionally, a model that is pre-trained on a diverse image task such as ResNet
can be transferred to our domain of texture analysis \footnote{All code is
    publicly available on GitHub (https://github.com/ngngardner/cuticulus)}.
However, a deep learning algorithm created specifically for the domain of
texture analysis had the best results in our experiments. In future work, we
will continue to explore texture classes present in the dataset which are not
captured by the binary class system. We will also investigate application of
texture-based deep maching learning in image texture analysis.

>>>>

\title{Investigation of ant cuticle dataset using image texture
    analysis}

@@

\address{%
    \addr{\affilnum{1}}{Laboratory of Machine Vision and Security Research, %
        College of Computing and Software Engineering, Kennesaw State %
        University, Marietta GA, USA}%
    \addr{\affilnum{2}}{College of Science and Mathematics, Kennesaw State
        University, Kennesaw, GA, USA}%
    \addr{\affilnum{3}}{Department of Mathematics, Kennesaw State University,
        Kennesaw, GA, USA}%
    \addr{\affilnum{4}}{Department of Informatics and Mathematics,
        Mercer University, Atlanta, GA, USA}%
}
\corraddr{chung1@kennesaw.edu}

@@

\begin{abstract}
    Ant cuticle texture presumably provides some type of function, and therefore
    is useful to research for ecological applications and bioinspired designs.
    In this study, we employ statistical image texture analysis and deep machine
    learning methods to classify similar ant species based on morphological
    features. We establish a public database of ant cuticle images for research.
    We provide a comparative study of the performance of image texture
    classification and deep machine learning methods on an ant cuticle dataset.
    Our results show that the deep learning methods give higher accuracy than
    statistical methods in recognizing ant cuticle textures. Our experiments
    also reveal that the deep learning networks designed for image texture
    performs better than the general deep learning networks.
\end{abstract}
\keywords{texture analysis; image processing; classification; machine learning;
    ant cuticle images; ecology}
\maketitle

@@

Insects compose half of biodiversity and rank among the most dominant organisms
in terrestrial ecosystems \cite{sheikh_diverse_2017}. A key factor for the
ecological success of insects is their exoskeleton, also known as cuticle. The
cuticle protects insects from predation, provides structural support, prevents
desiccation, and serves as a canvas for advertising visual and chemical signals
\cite{gullan_insects_2009}. Earlier research has focused on the macrostructures
and internal chemical components that describe the functionality of the
exoskeleton \cite{gunderson_insect_1989}. More recent work is being done to
understand the functional aspects of external cuticle micro sculpturing
\cite{muthukrishnan_insect_2020, watson_diversity_2017}.

Due to the extensive number of insect species, manual exploration of
insect-based information is difficult and often requires expert training.
Therefore, automated entomology has attracted biologists and computer
scientists, and is expected to be a major contribution to the future of
insect-based research \cite{martineau_survey_2017}. One of the most commonly
used data types for insect analysis is image data. To develop an image-based
system for insect analysis, we can utilize existing work in general
image analysis methods.

We study ants (\textit{Formicidae}) as they display an extreme diversity of
cuticle micro sculpturing across all subfamilies. Microsculpture ranges from
parallel longitudinal ridges to deep oval impressions to erratic protuberances.
Microsculpture has arisen convergently and independently throughout ant's
evolutionary history, which suggests  that these are complex traits undergoing
selection. Cuticle microsculpture on ants may help increase strength and
rigidness, resist abrasion, increase internal and external surface area, resist
microbial growth, and cultivate beneficial anti-biotic producing bacteria
\cite{johnson_effect_2011, bruckner_relationship_2017, currie_coevolved_2006}.
These specific functions may be associated with certain sculpturing types. To
analyze those functions, we first segment the image and group similar textures
for further analysis.

There is a large variety of ant species and most species are diverse in terms of
size, shape, behaviors, and cuticle textures \cite{harris_glossary_1979}. Based
on our initial observation of the cuticle sculpturing in ants, it appears that
some type of textural structures can be identified in images. Therefore, it is
worthwhile to explore image texture analysis for further knowledge discovery.
Image texture analysis has been used in image processing for many decades
\cite{hung_image_2019}. Due to the diversity of images on scales, smoothness,
coarseness, microtexture, and macrotexture occurring in images, it is a
difficult problem. With a rapid development of deep machine learning, deep
neural networks have been frequently used in image texture analysis in other
fields \cite{sun_robust_2014,liu_bow_2019,liu_pestnet_2019}.

Our contribution is to establish an ant image dataset which can be used to
explore the categorization of different ant cuticle textures using image texture
analysis methods. We document how the dataset of ant cuticle images is
established. Our experiment shows that the deep learning networks designed for
image texture analysis give a better accuracy than classical K-views clustering.

This paper is organized as below:  Section 2 reviews the microsculpturing
identification in ants, and image texture analysis and insect classification.
Section 3 describes the dataset preparation and image texture analysis methods
used in our experiments. Section 4 gives the experimental setup and methods
used. Section 5 shares visualization of both raw and processed images, and
analysis of experimental results. The conclusion then follows.

@@

One challenge for functional investigations of cuticle microsculpturing is the
complexity of terms used to describe these patterns. Complex terminology may be
useful for describing fine differences between closely related species, but
broader comparisons require a simplified classification scheme. We recently
developed a classification system for microsculpturing that consists of only
five categories: smooth, striate, punctate, reticulate, and tuberous
\cite{hellenbrand_ant_2022}. We used these categories to classify 11,722 ant
species, and found that the earliest ants were smooth and that cuticle
microsculpturing independently evolved later in multiple lineages
\cite{hellenbrand_ant_2022}. A simple classification scheme divides ants into
two categories: smooth and rough. Because microsculptuing has evolved numerous
times in ants, this binary classification scheme is sufficient.

The next challenge is to find out how to apply these schemes to large numbers of
species. Insect researchers have access to millions of images available through
digital museum collections \cite{beaman_mass_2012}. Going through these digital
collections manually is not feasible for large projects. There are over 14,000
described species of ants, and each species has multiple worker and reproductive
castes that may vary in their microsculpturing patterns. Even within an
individual ant, microsculpturing typically varies among and within body
segments. Machine learning offers a solution to this problem through developing
tools that automate trait classification. Automated approaches would allow
researchers to leverage existing image repositories to extract large trait
datasets that can then be used to detect evolutionary patterns.

@@

Pattern recognition methods have been developed for textural images based on the
spatial relationship of gray levels of pixels encoded in the texture
\cite{haralick_image_1985,hung_image_2019}. Those methods extract texture
features and the classification and segmentation methods are used for the
interpretation either using the supervised or unsupervised approaches
\cite{fukunaga_introduction_2013}. The interpretation method is usually modified
from the traditional pattern recognition methods. For example, the gray-level
co-occurrence matrix (GLCM) and local binary patterns (LBP) are two common
methods for extracting features \cite{haralick_textural_1973, goos_gray_2000}.

Classical image texture analysis methods can be grouped into four categories:
statistical, structural, model-based, and transform-based methods
\cite{bharati_image_2004}. The statistical method is often used by feature
extraction methods and then fed into the classification algorithm. Markov random
field models have also been used for textural image interpretation
\cite{hassner_use_1981, cross_markov_1983}. Transform-based methods use some
functions to decompose an image texture into a set of basic feature images.
Gabor filters and Wavelet expansions are two of the widely used approaches
\cite{bovik_multichannel_1990}.

\subsection{Deep learning}

Liu et al. gave a comprehensive survey on textural characterization in which
they call this feature extraction as texture representation \cite{liu_bow_2019}.
Their survey on bag of words and convolutional neural networks (CNNs)
\cite{krizhevsky_imagenet_2017} provides information on spatial relationship
concepts. The K-views was developed for taking the spatial information in the
clustering approach \cite{hung_image_2019}.

In the past decade, deep learning using CNNs has emerged as the mainstream
technology for image analysis. Following this trend, various related network
architectures have been designed to specifically characterize textural images.
Among these, the Fisher-vector CNN descriptor (FV-CNN) proposed by Cimpoi et al.
\cite{cimpoi_deep_2015} is widely accepted as one of the most important
pioneering works. It applies FV pooling to deep features obtained via a CNN
pre-trained using the ImageNet \cite{krizhevsky_imagenet_2017} to obtain encoded
features for texture classification. FV-CNN is capable of achieving much
improved classification accuracy comparing to traditional hand-crafted texture
features. However, CNN-based deep learning is only used in its feature
extraction module. Its encoding and classification modules still use traditional
learning methods, thus not supporting an end-to-end deep learning. In the
context of texture image classification, an end-to-end deep learning model
refers to a neural network architecture which consists of all the modules
(feature extraction, encoding and classification) between the initial input data
and the final output classification result, with all the modules being trained
simultaneously.

To achieve an end-to-end learning, Zhang et al. \cite{zhang_deep_2017} proposed
the deep texture encoding network (DeepTEN), in which a novel texture encoding
layer is added to a standard CNN architecture. Then, Xue et al.
\cite{xue_deep_2018} constructed the deep encoding pooling network which
improves over DeepTEN by integrating local spatial characteristics into the
texture representation. Based on these two methods, Hu et al. \cite{hu_multi_2019}
further developed the multi-level texture encoding and representation (MuLTER)
network, which embeds a learnable encoding module at each convolutional layer so
that encoding is performed for both low-level and high-level features, yielding
a multi-level texture representation.

@@

\subsection{Classification of insects}

@@

Feng et al. \cite{feng_automated_2013} apply an automated system to classify
moth images based on semantic related visual attributes, which are defined as a
pattern on the moth wings. Feng et al. \cite{feng_automated_2013} use a custom
texture descriptor based on the combination of GLCM and \textit{scale-invariant
    feature transform} (SIFT) features \cite{gotlieb_texture_1990,
    lowe_distinctive_2004}. The method is used to classify 50 different moth species
across 8 families \cite{feng_automated_2013}. Their results
\cite{feng_automated_2013} suggest that traditional feature extraction
techniques for the semantic visual attributes of the moth wings are sufficient
for training a classifier to classify an image between 10 randomly selected moth
species.

Urteaga et al. \cite{urteaga_scorpions_2016} use machine learning methods in
order to classify images between two different scorpion species:
\textit{Centruroides limpidus} and \textit{Centruroides noxius}. After applying
background distinction based on dynamic color threshold, they extract features
from the separated scorpion image such as aspect ratio, rectangularity, and
compactness. They apply three different models to classify the image as one of
the species: artificial neural network, regression tree, and random forest
classifiers \cite{urteaga_scorpions_2016}. Their results show that after
background removal, characteristics from the entire body of the scorpion can be
used to create a binary classifier that can classify the image as one of the two
species.

@@

In this section, we describe the creation of the custom dataset used in this
research. We use ant head images from AntWeb \cite{perrichot_antweb_2012} and
define two categories for them based on the appearance of the cuticle texture:
\textit{rough} and \textit{smooth}. For each species used, there exists only one
ant head image. We also refer to these ant head images, which describe the pose
of the ant, as ant cuticule images, due to the clear cuticle texture present on
the head. Some randomly selected images from each category are shown in Figures
\ref{fig:smooth-cuticle-texture} and \ref{fig:rough-cuticle-texture}.

@@

Initially, the sculpture identification protocol had 5 subcategories of cuticle
texture: smooth, punctate, striate, reticulate, and tuberous. For simplicity, we
worked only with the two main categories, and the other 4 categories were
combined to the single rough category. The training set identifications were
reviewed together as a group by the assistants. Once training was complete,
assistants were assigned the same genera of ants to identify independently each
week. A weekly meeting was held to discuss identifications and assign new ones.
These identifications were collected in the master spreadsheet and the
identifications were assigned to individual ant species on a majority basis.

To collect the images from AntWeb\footnote{\url{https://www.antweb.org/}}, the
assistants followed the taxonomy information available in the master spreadsheet
to the appropriate AntWeb page. In many cases, there are multiple ant head
images of the same species, and occasionally there are multiple image resolution
available from a single image. To simplify the data collection process, the
assistants were instructed to download the first ant head image of the species
being identified in the highest resolution possible. Each image was named with
an identifier that corresponds with the row number in the master spreadsheet.
The same ant head images that were downloaded in the data collection phase were
the same ones used in the sculpture identification protocol. Ant species which
did not have any images of the head were excluded from the dataset.
Additionally, ant species which only had a head image of a queen ant were
excluded from the dataset.

Ant specimen images in AntWeb \cite{perrichot_antweb_2012} are created by
different photographers and therefore have different attributes, such as
environment, resolution, and lighting. In the ant head images, the ant head is
in the center of the image and the body is pointing away from the camera. The
focus of the ant head image is centered on the head, with the background and
image artifacts from the ant body typically blurred. In most ant head images,
there is a bar which indicates the scale of the image due to the variety in the
sizes of different ant species. In a few ant head images, there exists some text
denoting the specimen identifier and other information. In terms of texture,
some ant specimens are very old, so their head images have other abnormalities
such as cracks in the cuticle and the presence of dust.

The completed dataset\footnote{All code is publicly available on GitHub
    \url{https://github.com/ngngardner/cuticulus}} contains 2,499 images as a
4-dimensional array with shape of (samples, rows, columns, channels), of
which there are 3 channels extracted from the RGB images. 1072 samples of
rough textured ant cuticle textures comprise 43\% of the dataset. The
remaining 1427 samples of smooth textured ant cuticle textures comprise 57\%
of the dataset. Due to the variety of the ant head image attributes, we
apply simple preprocessing before the images are used in our model. We want
the images to have a uniform size for simplicity in our classification
process. Since the ant head images are typically centered in the image, we
apply a center crop to each image to create a square image of the same size.
Once the image is square, we resize each image to a fixed size of 256x256
pixels. We leave other discrepancies in the images untouched. A summary of
the resulting dataset is shared in Table \ref{tab:dataset-summary}.

@@

\newcommand{\segmentedsubwidth}{0.24\textwidth}
\begin{figure}
    \centering
    \begin{subfigure}{\segmentedsubwidth}
        \includegraphics[width=1\linewidth]{/home/ngardn10/projects/cuticulus/paper/assets/images/kviews/r103.png}
        \caption*{Rough texture}
    \end{subfigure}
    \begin{subfigure}{\segmentedsubwidth}
        \includegraphics[width=1\linewidth]{/home/ngardn10/projects/cuticulus/paper/assets/images/kviews/r111.png}
        \caption*{Rough texture}
    \end{subfigure}
    \begin{subfigure}{\segmentedsubwidth}
        \includegraphics[width=1\linewidth]{/home/ngardn10/projects/cuticulus/paper/assets/images/kviews/s105.png}
        \caption*{Smooth texture}
    \end{subfigure}
    \begin{subfigure}{\segmentedsubwidth}
        \includegraphics[width=1\linewidth]{/home/ngardn10/projects/cuticulus/paper/assets/images/kviews/s107.png}
        \caption*{Smooth texture}
    \end{subfigure}

@@

\FloatBarrier
\section{Conclusion}
Ant cuticle texture presumably has some function, but without the proper tools,
evaluating the function based on thousands of species is infeasible. We have
shown in this work that a deep learning approach and classical image texture
analysis methods can be used to automatically categorize ants based on their
cuticle texture, therefore supporting research on the evaluation of the function
in future work. Our categorization system is novel in the field of automated
insect identification due to the broad number of species captured by it.
Additionally, a model that is pre-trained on a diverse image task such as ResNet
can be transferred to our domain of texture analysis \footnote{All code is
    publicly available on GitHub \url{https://github.com/ngngardner/cuticulus}}.
However, a deep learning algorithm created specifically for the domain of
texture analysis had the best results in our experiments. In future work, we
will continue to explore texture classes present in the dataset which are not
captured by the binary class system. We will also investigate application of
texture-based deep maching learning in image texture analysis.
